Functional Pearl: \\ Programming with Negative and Fractional Types

### 2012-02-18 dual.tex

\title{Continuations via Negative Information Flow}
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
We show a deep symmetry between functions and delimited continuations,
values and continuations that arises in {{langRev}} in a manner that
is reminiscent of Filinksi's Symmetric \lcal
~\cite{Filinski:1989:DCI:648332.755574}. The symmetry arises by
extending {{langRev}} with a notion of additive duality over the
monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
Compact Closed Categories. The resulting dual types, which we denote
{{-b}}, have a time traveling ``backward information flow''
interpretation and allow for the encoding of higher-order function and
iteration via the construction of {{trace}} operators, thereby making
the extended language {{langRevEE}} a Turing-complete reversible
programming language with higher-order functions and first-class
delimited continuations.
\end{abstract}

@@

@@

\paragraph*{Ariola, Herbelin, and Sabry.} We use subtractive types. Things
are constrained so that the positive and negative values must be used
together. In our new work we have no such constrained, the negative value can
flow anywhere. The entire system guarantees that any closed program would
have to account for it. We don't have to introduce special constraints to
achieve that.

\paragraph*{Focusing, Polarization, etc.}

>>>>

\title{Functional Pearl: \\ Programming with Negative and Fractional Types} 
\authorinfo{Roshan P. James}
           {Indiana University}
           {rpjames@indiana.edu}
\authorinfo{Amr Sabry}
           {Indiana University}
           {sabry@indiana.edu}
\maketitle

\begin{abstract}
  Every functional programmer knows about sum and product types, $a+b$ and
  $a*b$ respectively. Negative and fractional types, $a-b$ and $a/b$
  respectively, are much less known and their computational interpretation is
  unfamiliar and often complicated. We show that in a programming model in
  which information is preserved (such as the model introduced in our recent
  paper on \emph{Information Effects}), negative and fractional types have
  particularly intuitive and natural computational
  interpretations. Intuitively, values of negative types are values that flow
  ``backwards'' to satisfy demands, and values of fractional types are values
  that represent first-class pattern clauses. The combination of negative and
  fractional types allow a programmer to express a wide range of programming
  idioms, including higher-order functions, delimited continuations,
  speculative computation, logic programming, and more. 
\end{abstract}

@@

We show a deep symmetry between functions and delimited continuations,
values and continuations that arises in {{langRev}} in a manner that
is reminiscent of Filinksi's Symmetric \lcal
~\cite{Filinski:1989:DCI:648332.755574}. The symmetry arises by
extending {{langRev}} with a notion of additive duality over the
monoid {{(+, 0)}} by including {{eta}} and {{eps}} operators of
Compact Closed Categories. The resulting dual types, which we denote
{{-b}}, have a time traveling ``backward information flow''
interpretation and allow for the encoding of higher-order function and
iteration via the construction of {{trace}} operators, thereby making
the extended language {{langRevEE}} a Turing-complete reversible
programming language with higher-order functions and first-class
delimited continuations.

@@

\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
subtraction type because its ``computational interpretation is not
familiar.'' Curien and Herbelin study duality in classical logic, show that
it exchanges call-by-value and call-by-name. They extend classical natural
deduction with subtraction but give it no computational meaning. Crolard (in
the formulae-as-types interpretation of subtractive logic) address the
computational interpretation of substraction. He explains the type $A-B$ as
the of \emph{coroutines} with a local environment of type $A$ and a
continuation of type $B$. The description is complicated by what is
essentially the desire to enforce linearity constraints so that coroutines
cannot access the local environment of other coroutines. Must cite Selinger
control categories in this context of duality but I am not sure what to say:
it assumes cartesian closed categories for one thing and not symmetric
monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
explain delimited continuations.

\paragraph*{Linear Logic} In accounts that are linear, the value and
continuation that comprise the substractive type need to be constrained to
``stay together.'' This can be achieved by various restrictions. In this work
we have no such constraints, the negative value can flow anywhere. The entire
system guarantees that any closed program would have to account for it. We
don't have to introduce special constraints to achieve that. Zeilberger in
the paper on polarity and the logic of delimited continuations uses polarized
logic: he shows that if positive and negative values are completely symmetric
except that answer types are positive, then the framework accommodates
delimited continuations. But he interprets negative values are control
operators, or as values defined by the shape of their continuations. We
simply interpret values of negative type as values flowing in the ``other''
direction.
